A Dearth of Young Stars and Planets in the Kepler Field

### 2024-05-21 paper/ms.tex
update ms

Rotation-based ages have been reported for
various subsets of Kepler stars since the early data releases
\citep[e.g.][]{Walkowicz_2013,McQuillan_2014,Reinhold_2015,Angus_2018}.
More recent work has further explored incorporating information from
stellar kinematics \citep{2021AJ....161..189L,2024AJ....167..159L},
and from light curve features such as starspot amplitudes
\citep{2023ApJ...952..131M}.  Our analysis is motivated by a few
factors that can yield improvements, particularly for the ages of
sub-gigayear stars.   These factors are as follows.
\begin{enumerate}[label={\it \roman*)},leftmargin=12pt,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has now reached maturity
    \citep[e.g.][]{Thompson_2018};
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
    but also how the astrophysical dispersion of stars around this average
    converges by the $\approx$700\,Myr age of Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item We can fit the open cluster data, and use these fits to infer
    ages by marginalizing over the range of ages that might explain
    any one star's rotation period \citep{Bouma_2023}.
    This represents an improvement in both accuracy and in realism
    of uncertainty propagation relative to previous calibrations.
  %
  \item We can identify binary stars with greater fidelity than
    in the past, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
    These clusters offer a sample of stars with which we can test the
    reliability of single-star age-dating methods.
\end{enumerate}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.98\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf Literature ages vs.~benchmark open cluster ages.}
		Each point denotes a reported age,
		normalized by the age of a stellar ensemble in which the star
		is a candidate member.
		We used candidate cluster members from previous studies that used spatial and kinematic
		information.  Although some field interlopers may be present in the 
		membership lists,
		any outliers can be compared between different methods on a
		relative basis.
		The number of stars in any given cluster follows by crossmatching every study against
		a fixed cluster list; certain methods report ages for more stars than others.
		Horizontal scatter is artificially added to visually
		clarify the statistical age uncertainties.
	}
	\label{fig:agescalecompone}
\end{figure*}

@@

\begin{figure*}[!t]
  \begin{center}
    \leavevmode
    \subfloat{
    \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
    }

	\vspace{-0.35cm}
    \subfloat{
    \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
    }
  \end{center}
  \vspace{-0.6cm}
  \caption{{\bf Kepler's demographic cliff,} visible in the
  rotation-derived ages of stars (left) and planet hosts (middle) in
  the Kepler field.  The top row corresponds to all stars with
  temperatures of 3800--6200\,K for which we calculated rotation-based
  ages.  The darker lines impose quality cuts on factors such as
  binarity, crowding, and the star's evolutionary state.  The bottom
  row selects only stars with temperatures of 4400-5400\,K, which have
  relatively precise ages near 2\,Gyr due to their fast spin-down
  \citep[see][]{Bouma_2023}.  The statistical uncertainty for a mean
  individual star at 1, 2, and 3\,Gyr is shown, and is identical
  across each row.  Finally, the right panel compares the
  rotation-based age distributions against star formation histories
  derived using CMD fitting.  Completeness is near unity until
  $t$$\lesssim$3\,Gyr \citep{2022ApJ...937...94M}.
  \label{fig:hist_tgyro}
  }
\end{figure*}

@@

{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex currently has its
most constraining age measurement from relative isochrone dating of
pre-main-sequence M dwarfs; it contains Kepler-1627Ab, Kepler-1643b,
Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022a,Bouma_2022b}.
Different sub-groups of the complex may vary in age by $\approx$50\%
(\citealt{Bouma_2022b}; Kerr~et al.,~submitted).  The planets
themselves are all on close-in orbits (5-25\,days), with sizes from
2-4\,$R_\oplus$.  The rotation, lithium, and cluster ages all agree
for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the cluster age reported by \citet{Bouma_2022b} for the RSG-5
subgroup.

{\it MELANGE-3}---\citet{Barber_2022} reported a
$105\pm10$\,Myr association in the Kepler field and derived its age
using rotation, lithium, and the isochronal pre-main-sequence.
Kepler-970 and Kepler-1928 harbor two transiting mini-Neptunes inside
this association.  Our rotation-based
age-dating independently recovers these systems, with $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and Kepler-1928 respectively.  The former rotation-based
age being slightly older than the suggested Pleiades age for the
assocation agrees with Figure~4 of \citet{Barber_2022}.

{\it Theia~520}---We derive rotation-based ages of 330-350\,Myr for
Kepler-968 and Kepler-52.  These stars were reported by
\citet{2019AJ....158..122K} to be members of an association called
Theia~520 based on their spatial and kinematic correlations with
nearby stars.  The core of this structure is also known as UBC-1
\citep{2018A&A...618A..59C}; since both Kepler planet hosts are in
diffuse regions of the cluster, we use the name from
\citeauthor{2019AJ....158..122K}.  Based on isochrone-fitting,
Theia~520 is $247^{+64}_{-51}$\,Myr old \citep{2019AJ....158..122K}.
An independent analysis by \citet{2024A&A...681A..13F} using g-mode
asteroseismology, gyrochronology, and isochrone fitting found
consistent results.  Ongoing work by J.~Curtis et al.~broadly finds
that the two planet-hosting stars are indeed members of the diffuse
population.  Given the cluster-level and individual-star evidence,
this makes Kepler-968 and Kepler-52 the youngest multiplanet systems
currently known from the primary Kepler mission.

{\it NGC~6811}---\citet{Meibom_2013} reported the discovery of
Kepler-66 and -67b, two mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).
We report rotation-based ages for these systems of
\kepsixsixtgyro\,Myr and \kepsixseventgyro\,Myr for Kepler-66 and
-67b, respectively.  Kepler-66 ($T_{\rm eff}$$\approx$5900\,K, $P_{\rm
rot}$$\approx$10.4\,days) has a lower precision and a more asymmetric
posterior because the spindown rates of stars this hot slow down after
the first gigayear.  Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
rot}$$\approx$10.3\,days) is marginally hotter than the later K dwarfs
that have ``stalled'' spindown from 0.7 to 1\,Gyr, enabling its
rotation period to be diagnostic of the stellar age.

@@

We compared the independent rotation and lithium-based ages of the
known planet hosts using the point-estimates of the available
posteriors.  There are three cases of interest:

@@

Appendix~\ref{app:inconsistent} describes all cases of ``discrepant''
confirmed planets for which a sub-gigayear age is reported by at least
one age indicator.  All but two systems are either evolved stars or
else unresolved binaries, and are automatically flagged as such.  The
most interesting exception is Kepler-786,
an early K dwarf with $t_{\rm
Li}$=\kepseveneightsix\,Myr, which would imply that rotation should be
detectable.  However no photometric rotation signal is detected; other
age indicators in the spectrum similarly suggest that the star may not
in fact be young.

@@

Our approach to answering these questions was to curate a sample of
stellar rotation periods, lithium equivalent widths, and temperatures
using archival and new data.  We derived new ages using empirical
methods that have been calibrated against an extensive set of open
cluster data.

Our analysis blindly recovered all 14 known Kepler planets in
clusters, with ages in agreement with the more precise cluster-based
ages.  Although lithium provided only marginal added information for
most of the sample, the independent lithium and rotation ages agreed
for $\gtrsim$90\% of cases.  In particular, we reported {\bf XX}
confirmed $<$500\,Myr planets with strong age constraints from both
rotation and lithium, including {\bf YY} mini-Neptunes, {\bf ZZ}
super-Earths, {\bf AA} Earths and sub-Earths, and a Neptune-sized
planet.

>>>>

Rotation-based ages have been reported for
various subsets of Kepler stars since the early data releases
\citep[e.g.][]{Walkowicz_2013,McQuillan_2014,Reinhold_2015,Angus_2018}.
More recent work has further explored incorporating information from
stellar kinematics \citep{2021AJ....161..189L,2024AJ....167..159L},
and from light curve features such as starspot amplitudes
\citep{2023ApJ...952..131M}.  Our analysis is motivated by a few
factors that can yield improvements, particularly for the ages of
sub-gigayear stars.   These factors are as follows.

\begin{enumerate}[label={\it \roman*)},leftmargin=12pt,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has now reached maturity
    \citep[e.g.][]{Thompson_2018}.
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
    but also how the astrophysical dispersion of stars around this average
    converges by the $\approx$700\,Myr age of Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B}.
  %
  \item We can fit the open cluster data, and use these fits to infer
    ages by marginalizing over the range of ages that might explain
    any one star's rotation period \citep{Bouma_2023}.
    This represents an improvement in both accuracy and in realism
    of uncertainty propagation relative to previous calibrations.
  %
  \item We can identify binary stars with greater fidelity than
    in the past, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
    These clusters offer a sample of stars with which we can test the
    reliability of single-star age-dating methods.
\end{enumerate}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.99\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf Literature ages vs.~benchmark open cluster ages.}
		Each point denotes a reported age,
		normalized by the age of a stellar ensemble in which the star
		is a candidate member.
		We used candidate cluster members from previous studies that used spatial and kinematic
		information.  Although some field interlopers may be present in the 
		membership lists,
		any outliers can be compared between different methods on a
		relative basis.
		The number of stars in any given cluster follows by crossmatching every study against
		a fixed cluster list; certain methods report ages for more stars than others.
		Horizontal scatter is artificially added to visually
		clarify the statistical age uncertainties.
	}
	\label{fig:agescalecompone}
\end{figure*}

@@

\begin{figure*}[!t]
  \begin{center}
    \leavevmode
    \subfloat{
    \includegraphics[width=.99\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
    }

	\vspace{-0.35cm}
    \subfloat{
    \includegraphics[width=.99\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
    }
  \end{center}
  \vspace{-0.66cm}
  \caption{{\bf Kepler's demographic cliff,} visible in the
  rotation-derived ages of stars (left) and planet hosts (middle) in
  the Kepler field.  The top row corresponds to all stars with
  temperatures of 3800--6200\,K for which we calculated rotation-based
  ages.  The opaque lines impose quality cuts on factors such as
  binarity, crowding, and the star's evolutionary state.  The bottom
  row selects only stars with temperatures of 4400-5400\,K, which have
  relatively precise ages near 2\,Gyr due to their fast spin-down.
  The statistical uncertainty for a mean
  individual star at 1, 2, and 3\,Gyr is shown, and is identical
  across each row.  Finally, the right panel compares the
  rotation-derived ages against star formation histories
  derived using CMD fitting.  Completeness in Kepler's
  $P$$_{\rm rot}$ detection sensitivity is near unity until
  $t$$\lesssim$3\,Gyr \citep{2022ApJ...937...94M}.
  \label{fig:hist_tgyro}
  }
\end{figure*}

@@

{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex has been
dated using the isochrone loci of its
pre-main-sequence M dwarfs; it contains Kepler-1627Ab, Kepler-1643b,
Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022b,Bouma_2022a}.
Different sub-groups of the complex may vary in age by $\approx$50\%
(\citealt{Bouma_2022b}; Kerr~et al.,~submitted).  The planets
themselves are all on close-in orbits (5-25\,days), with sizes from
2-4\,$R_\oplus$.  The rotation, lithium, and cluster ages all agree
for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the cluster age reported in \citet{Bouma_2022b} for the RSG-5
subgroup.

{\it MELANGE-3}---\citet{Barber_2022} reported a
$105\pm10$\,Myr association in the Kepler field and derived its age
using rotation, lithium, and the isochronal pre-main-sequence.
Kepler-970 and Kepler-1928 harbor two transiting mini-Neptunes inside
this association.  We find rotation-based ages
of $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and Kepler-1928 respectively.  The former rotation-based
age being slightly older than the suggested Pleiades age for the
assocation agrees with Figure~4 of \citet{Barber_2022}.

{\it Theia~520}---We derive rotation-based ages of 330-350\,Myr for
Kepler-968 and Kepler-52.  These stars were reported by
\citet{2019AJ....158..122K} to be members of an association called
Theia~520 based on their spatial and kinematic correlations with
nearby stars.  The core of this structure is also known as UBC-1
\citep{2018A&A...618A..59C}.  Based on isochrone-fitting,
Theia~520 is $247^{+64}_{-51}$\,Myr old \citep{2019AJ....158..122K}.
An independent analysis by \citet{2024A&A...681A..13F} using g-mode
asteroseismology, gyrochronology, and isochrone fitting found
consistent results.  Ongoing work by \citet{Curtis2024} broadly finds
that the two planet-hosting stars are indeed members of the diffuse
population.  Given the cluster-level and individual-star evidence,
this makes Kepler-968 and Kepler-52 the youngest multiplanet systems
currently known from the main Kepler mission.

{\it NGC~6811}---\citet{Meibom_2013} reported the discovery of
Kepler-66 and -67b, two mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).
We report rotation-based ages for these systems of
\kepsixsixtgyro\,Myr and \kepsixseventgyro\,Myr for Kepler-66 and
-67b, respectively.  Kepler-66 ($T_{\rm eff}$$\approx$5900\,K, $P_{\rm
rot}$$\approx$10.4\,days) has a lower precision and a more asymmetric
posterior because of its slow spindown rate.  Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
rot}$$\approx$10.3\,days) is marginally hotter than the later K dwarfs
that have ``stalled'' spindown from 0.7 to 1\,Gyr, enabling its
rotation period to be diagnostic of the stellar age.

@@

We compared the independent rotation and lithium ages of the
known planet hosts using the point-estimates of the available
posteriors.  There are three cases of interest:

@@

Appendix~\ref{app:inconsistent} describes all cases of ``discrepant''
confirmed planets for which a sub-gigayear age is reported by at least
one age indicator.  All but two systems are either evolved stars or
else unresolved binaries, and are automatically flagged as such.  The
more interesting of the two is Kepler-786,
an early K dwarf with $t_{\rm
Li}$=\kepseveneightsix\,Myr, which would imply that rotation should be
detectable.  However no photometric rotation signal is detected; other
age indicators in the spectrum similarly suggest that the star may not
in fact be young.

@@

Our approach to answering these questions was to curate a sample of
stellar rotation periods, lithium equivalent widths, and temperatures
using archival and new data.  We derived new ages using empirical
interpolation-based methods, and assessed the reliability of these ages
by comparing them against benchmark open clusters
(Figure~\ref{fig:agescalecompone}).

Our analysis blindly recovered the youth of all 14 known Kepler planets in
clusters, with ages that agreed with the more precise ensemble ages.
While lithium provided only marginal added information for
most of the sample, in cases for which comparison was possible, the independent lithium and rotation ages agreed
$\gtrsim$90\% of the time.  In particular, we reported {\bf XX}
confirmed $<$500\,Myr planets with strong age constraints from both
rotation and lithium, including {\bf YY} mini-Neptunes, {\bf ZZ}
super-Earths, {\bf AA} Earths and sub-Earths, and a Neptune-sized
planet.
